---
title: "Cleaning US Influenza Admissions Data"
subtitle: "From a raw surveillance CSV to a tidy, forecast-ready table"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,        # show the code so students can read along
  message = FALSE,    # hide package startup chatter
  warning = FALSE     # keep the narrative clean; we validate explicitly below
)
```

## What this script does, and why

Before we can *explore* or *model* respiratory data, we need a clean, predictable table to build on. Raw surveillance exports are rarely in the shape we want: they carry columns we don't need, mix geographies together, and store numbers as text.

This document walks through a small cleaning pipeline that takes the CDC weekly Hospital Respiratory Data (HRD) export and produces one tidy file: **one row per week** of **US** influenza admissions, with exactly three columns — `week`, `location`, and `value`. That three-column shape is deliberate: it matches the tidy layout the rest of our modeling and FluSight work expects downstream.

A theme to notice throughout: **we make our assumptions explicit and check them.** Nearly every block below either states what it expects or verifies that it got it. That habit is what turns a script you *hope* works into a script you can *trust*.

## Packages

We only need two packages here. `readr` gives us fast, predictable CSV reading and writing with explicit control over column types. `dplyr` gives us the verbs (`filter`, `transmute`, `arrange`) that make the transformation readable.

```{r packages}
library(readr)
library(dplyr)
```

## Define file paths in one place

We name every file the script touches up front, rather than scattering paths through the code. This gives us a single source of truth: if the data moves or we rename an output, we change it once, here.

Notice the structure of the output paths — `output/data/01_cleaning/...` and `output/figures/01_cleaning/...`. Keeping each script's outputs in a folder named after the script (`01_cleaning`) means results never collide and it's always obvious which step produced which file.

```{r paths}
input_path  <- "data/Weekly Hospital Respiratory Data (HRD) Metrics by Jurisdiction.csv"
output_path <- "output/data/01_cleaning/cleaned_flu_admissions.csv"
figure_path <- "output/figures/01_cleaning/epicurve_us_flu_admissions.png"
```

> **Knitting note:** these are *relative* paths, so they assume your working directory is the project root (the folder that contains `data/`). When you knit an `.Rmd`, R uses the document's own folder as the working directory by default, so if the knit can't find the input file, that's usually the reason.

## Create output folders if they don't exist

A script shouldn't assume its destination folders are already there. `dir.create` with `recursive = TRUE` builds the full folder tree, and `showWarnings = FALSE` keeps it quiet when the folders already exist (which is the normal case on a re-run). `dirname()` just strips the filename off, leaving the folder to create.

```{r make-dirs}
dir.create(dirname(output_path), recursive = TRUE, showWarnings = FALSE)
dir.create(dirname(figure_path), recursive = TRUE, showWarnings = FALSE)
```

## Read only the columns we need

The raw export is wide, but we only care about three things: the week, the geography, and the influenza admissions count. `col_select` pulls just those, which is faster and keeps our attention on what matters.

Two deliberate choices are worth calling out:

-   **Read everything as character** (`col_types = cols(.default = col_character())`). It may feel odd to read numbers as text, but it means *we* decide when and how each column becomes a number — instead of letting `readr` guess from the first few rows and occasionally get it wrong on messy surveillance data. We convert on purpose in the next step.
-   **`any_of(...)` for the admissions column.** Different exports of this dataset spell the column either `Total.Influenza.Admissions` (dots) or `Total Influenza Admissions` (spaces). `any_of()` will grab whichever one is present and silently ignore the other, so the read doesn't break either way.

```{r read-raw}
raw <- read_csv(
  input_path,
  col_select = c(
    `Week Ending Date`,
    `Geographic aggregation`,
    any_of(c("Total.Influenza.Admissions", "Total Influenza Admissions"))
  ),
  col_types = cols(.default = col_character()),
  show_col_types = FALSE
)
```

## Figure out which spelling of the column we actually got

We handled *reading* either spelling above; now we need to know which one landed in `raw` so we can refer to it downstream. We check both candidate names against the columns that actually exist and keep the first match.

If neither is present, something is wrong with the input file, and we `stop()` immediately with a clear message rather than failing mysteriously three steps later. **Fail early, fail loudly, and say why** — future-you (or a classmate re-running this) will thank you.

```{r resolve-column}
# Support both dotted and spaced versions of the influenza admissions column name.
influenza_col <- c("Total.Influenza.Admissions", "Total Influenza Admissions")
influenza_col <- influenza_col[influenza_col %in% names(raw)][1]

if (is.na(influenza_col)) {
  stop("Could not find influenza admissions column in input data.")
}
```

## Reshape into the tidy table we want

This is the heart of the script. Reading it as a pipeline, top to bottom:

-   **`filter(...)`** keeps only the national ("USA") rows, dropping every state/territory. This step is why our output is US-only.
-   **`transmute(...)`** builds a brand-new set of columns and drops everything else (unlike `mutate`, which keeps the originals). We create exactly the three we want:
    -   `week` — the `Week Ending Date` text turned into a real `Date`, so it sorts and plots correctly as time.
    -   `location` — hard-coded to `"US"`, a stable label for the national series.
    -   `value` — the admissions count. `parse_number()` pulls the numeric value out of the character column, tolerating stray commas or symbols. `.data[[...]]` lets us reference the column by the *name we resolved* a moment ago, whichever spelling it turned out to be.
-   **`arrange(week)`** sorts chronologically, so the series reads left-to-right in time — important for both the checks and the plot that follow.

```{r clean}
cleaned <- raw %>%
  filter(`Geographic aggregation` == "USA") %>%
  transmute(
    week = as.Date(`Week Ending Date`),
    location = "US",
    value = readr::parse_number(.data[[influenza_col]])
  ) %>%
  arrange(week)
```

## Take a look at the cleaned data

Before trusting anything, let's actually *see* what we built. `dim()` reports the shape — how many weeks (rows) and columns we ended up with — and `head()` prints the first several rows so we can eyeball the values. Notice the column types shown under each name: `week` is a `date`, `location` is text (`chr`), and `value` is a number (`dbl`) — exactly the three-column, correctly-typed layout we were aiming for.

```{r peek}
dim(cleaned)     # number of weeks (rows) and columns
head(cleaned, 10) # first ten weeks of the cleaned series
```

## Check that the result is what we promised

Here we verify the cleaned table against a small contract we set for ourselves. `stopifnot()` halts the script the moment any condition is `FALSE`, so a bad result can never quietly slip into the next stage of the pipeline.

Read these as a checklist of everything we're claiming about `cleaned`:

-   there's at least one row (the filter didn't wipe everything out);
-   the columns are *exactly* `week`, `location`, `value`, in that order;
-   every row is US;
-   `week` really is a `Date` and `value` really is numeric;
-   there are no missing values hiding in `value`.

Writing checks like these is a genuinely useful habit: they're an objective, machine-verifiable definition of "done" that the data has to satisfy before any later step is allowed to build on it.

```{r checks}
# Acceptance checks from the agent instructions.
stopifnot(nrow(cleaned) > 0)
stopifnot(all(names(cleaned) == c("week", "location", "value")))
stopifnot(all(cleaned$location == "US"))
stopifnot(inherits(cleaned$week, "Date"))
stopifnot(is.numeric(cleaned$value))
stopifnot(!anyNA(cleaned$value))
```

## Save the cleaned data

With the table validated, we write it out as a CSV — the artifact the next scripts will read. The final `stopifnot(file.exists(...))` confirms the file actually landed on disk, closing the loop on "did this step really succeed?"

```{r write}
write_csv(cleaned, output_path)
stopifnot(file.exists(output_path))
```

## Draw an epicurve as a visual sanity check

Numbers passing checks is one kind of confidence; *seeing* the data is another. A quick epidemic curve (an "epicurve") — admissions plotted over time — instantly reveals whether the series has the seasonal flu shape we'd expect, or whether something looks off.

We'll do two things with this plot: **save a high-resolution copy to disk** (the version the pipeline and any report would use) and **render it inline** so you can see it right here in the knitted document.

First, the saved-to-disk version. A few details in this plotting block:

-   `png(...)` opens a file to draw into, so the chart is **saved to disk** rather than shown inline. `res = 150` gives a crisp image at a readable size.
-   `par(no.readonly = TRUE)` saves the current graphics settings, and `on.exit(par(op), add = TRUE)` restores them afterward — good manners so this block doesn't leave the plotting device in a weird state.
-   `barplot(...)` draws one bar per week. `space = 0` and `border = NA` pack the bars into a continuous curve; `las = 2` rotates the date labels vertical and `cex.names = 0.5` shrinks them so many weeks fit without overlapping.
-   `dev.off()` closes the file (this is what actually finishes writing the PNG), and the final check confirms the figure exists.

```{r epicurve}
png(filename = figure_path, width = 1400, height = 800, res = 150)
op <- par(no.readonly = TRUE)
on.exit(par(op), add = TRUE)

plot_values <- as.numeric(cleaned$value)

barplot(
  height = plot_values,
  names.arg = format(cleaned$week, "%Y-%m-%d"),
  space = 0,
  border = NA,
  col = "#1f78b4",
  las = 2,
  cex.names = 0.5,
  main = "US Weekly Influenza Admissions Epicurve",
  xlab = "Week Ending Date",
  ylab = "Admissions"
)

dev.off()
stopifnot(file.exists(figure_path))
```

### The same curve, rendered inline

Here we draw the epicurve straight into the document. There's no `png()` around it, so R Markdown catches the plot and embeds it in the knitted output. The chunk options `fig.width` and `fig.height` control its size on the page, and `fig.cap` adds a caption underneath.

We can reuse `plot_values` from the block above, since chunks in a knitted document share one environment and run top to bottom. (If you find yourself copying the `barplot()` call around a lot, that's a good nudge to wrap it in a small helper function and call it in both places — the classic "don't repeat yourself" move.)

```{r epicurve-inline, fig.width=12, fig.height=6, fig.cap="US weekly influenza admissions."}
barplot(
  height = plot_values,
  names.arg = format(cleaned$week, "%Y-%m-%d"),
  space = 0,
  border = NA,
  col = "#1f78b4",
  las = 2,
  cex.names = 0.5,
  main = "US Weekly Influenza Admissions Epicurve",
  xlab = "Week Ending Date",
  ylab = "Admissions"
)
```

## Where this leads

We now have a small, trustworthy file — `cleaned_flu_admissions.csv` — plus a picture of the series. That clean table is the foundation for the next steps: exploring the data more deeply and, from there, fitting forecasting models. Every minute spent making this step reliable pays off in every step that builds on it.
